The Evolution of Phenotypically Invariant Genetic Networks

### 2021-04-14 review-responses.tex

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are pleased to submit a revision of our manuscript, ``{System drift and speciation},'' for your review. 
We apologize for the delayed resubmission but hope that you will find the
revised manuscript significantly improved.

We thank you and the two reviewers for providing constructive suggestions 
and feedback that has helped to improve the manuscript. We have followed 
the suggestions closely and hope to have fully addressed all the concerns and 
suggestions as detailed in our point-by-point response below.

@@

In our original submission, we studied the effect of gene regulatory network evolution on speciation,
first by applying linear system theory to characterize the large space of phenotypically equivalent network organizations,
and second by using quantitative genetics to show that neutral genetic drift on this network space can lead to speciation over plausible timescales.

We agree with the reviewers that the combination of these two approaches, while interesting,
was not seamless and that the quantitative genetics section required further development, 
and thus will follow the recommendation of both the Editor and Associate Editor, to break the manuscript into
two papers. One paper will focus on quantitative genetics, and the other -- 
the focus of this present revision -- will focus on system theory. 
A revised manuscript focusing on quantitative genetics will be submitted at a later date.

We hope that by focusing this paper on system theory, we have clarified our results, which include a mathematical description of phenotypically equivalent network space 
and its implications for speciation under neutral genetic drift. These results suggest that the space of phenotypically equivalent network organizations is substantial,
and often not reproductively compatible, leaving many opportunities for independently evolving populations to become reproductively incompatible.
The linear system framework applied in this manuscript explicitly describes possible molecular pathways and naturally predicts Haldane's rule.

We would like to note that the previous version of this paper as a preprint has since received 10 citations as well as attracted
substantial positive attention as a preprint (e.g., see tweets at
\url{https://www.biorxiv.org/content/10.1101/231209v2}). Furthermore, this paper has been downloaded more than 2,229 times (\url{https://rxivist.org/papers/24911}),
making it the 156$^{\text{th}}$ most downloaded evolutionary biology paper available on the \emph{bioRxiv}, as of this writing.

### 2021-04-23 review-responses.tex

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are writing to submit a revision of our manuscript, ``{System drift and speciation},'' for your review. 
We sincerely apologize for the delayed resubmission but hope that you will find the
revised manuscript significantly improved.

@@

We thank you and the two reviewers for providing constructive suggestions 
and feedback that has helped to improve the manuscript. We have followed 
the suggestions closely and hope to have fully addressed all the concerns and 
suggestions as detailed in our point-by-point response below.

In our original submission, we studied the effect of gene regulatory network evolution on speciation,
first by applying linear system theory to characterize the large space of phenotypically equivalent network organizations,
and second by using quantitative genetics to show that neutral genetic drift on this network space can lead to speciation over plausible timescales.

We agree with the reviewers that the combination of these two approaches, while interesting,
was not seamless and that the quantitative genetics section required further development, 
and thus have followed the recommendation of both the Editor and Associate Editor,
by breaking the manuscript into two papers.
One paper -- the present revision -- covers the application of system theory to regulatory networks,
and the other (which we plan to resubmit soon) will cover the quantitative genetics.
In addition to detailed responses to the reviews,
we are submitting a color-coded diff to make it easy to see what has been removed.
We are happy with the result -- as predicted by the reviewers, the resulting paper stands alone
without the quantitative genetics results.

### 2021-04-27 review-responses.tex

\vskip 4em

\pagebreak

%%%%%%%%%%%%%%
\reviewersection{AE}

\begin{quote}
I have received the evaluations of two reviewers, and I have read the paper
myself. First, sorry for the long time it took to review your manuscript; it
was reviewed on the timescale of mathematicians (for Reviewer2), but both
reviewers provided thorough evaluations of your work (which in my opinion is
better than a quick but superficial review).
\end{quote}

Thanks for the apology,
but given the delay in our resubmission, we can hardly complain, at this point.

\begin{quote}
Your manuscript is quite long already, and the reviewers' suggestions of
modifications and clarification, which need to be implemented, will make the
manuscript even longer. I however share the reviewers' opinion that the
manuscript's two parts are only loosely related. My suggestion is therefore to
publish them separately. Although connecting the systems biology and the
quantitative genetics parts is indeed an exciting endeavor, this is not really
achieved in the current version of the manuscript, and it would be more
profitable to first better describe each part separately. Regarding the
connexion between the two parts, please also pay particular attention to R2's
first specific comment about dimensionality.

@@

This may be because I am more familiar with Fisher's geometric model than
systems biology models, but it seems to me that the first part is more novel,
and should therefore deserve your attention first, should you follow my
suggestion of publishing the two parts separately.
\end{quote}

Thanks for the careful attention to the paper.
We have decided to go with this suggestion,
and have split the paper in two;
this paper is the ``systems biology'' part of it.
We've left in reviewer comments regarding the ``Fisher's geometric model'' portion below,
commenting that the relevant bit ``has been removed''.

@@

\begin{point}{p5, Note on implementation}
    ``the closure of span(B)'' may not be
    understandable by most of Evolution's readers; please explain what this means.
\end{point}

\reply{
    We've added a clarifying note \revref{}
    but we're hoping that readers not familiar with linear algebra will realize
    that they can skip this ``note on implementation'' without loss of continuity.
}

\begin{point}{Figures 1-3:}
    In figures 1 and 3, there are two types of arrow heads (+ and -
effects), but not in Figure 2, probably because the sign of the effect depends
on the value of $\tau$. Maybe use a  different type of arrow heads for
interactions that may change signs, to avoid the confusion with positive
effects?
\end{point}

\reply{
  We think that adding a third type of arrowhead may only add to the confusion, so now we are using only the standard arrowhead.
}

@@

\reply{
    This part has been removed.
}

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
The ms combines two parts that are only loosely linked: one ``systems biology''
part and a ``popgen'' part. The combination of both parts is appealing, primarily
because we lack such combined approaches and it is not easy to come up with an
adequate framework. Below, I will discuss both parts in turn.
\end{quote}

This manuscript is mostly the ``systems biology'' part,
but we think that the potential applications to ``popgen'' are still reasonably visible.
(And, we plan to resubmit the ``popgen'' part separately.)

% \begin{quote}
%     \textbf{(I)} The first (systems biology) part makes use of the concept of
% ``nonidentifiability'' of chemical reaction networks and develops these concepts
% in the context of a genotype to phenotype map. In particular, this map is
% analyzed for the case where it can be modeled as a linear system.
% \end{quote}

@@

\begin{point}{}
    First note that the concept of ``nonidentifiability'' is closely related to
themes that have been discussed in evolutionary biology for a long time, under
names like ``mutational robustness'', ``network neutrality'', ``canalization'',
``redundancy'' etc. In particular, if you characterize your model by ``many
distinct (and mutationally connected) molecular pathways can realize identical
phenotypes'', note that this essentially describes what has been called a
``neutral network'' and studied under this name in many articles. Also concrete
developmental networks have been studied in this context (eg, von Dassow etal
2000, Nature 406:188–192). Discussion of this literature - and more importantly
discussion of terms and notions used here relative to terms that have been used
elsewhere - is largely missing. I won't be able to summarize all this here, but
an older review is the Evolution Perspective piece ``Evolution and detection of
genetic robustness'' (de Visser et al 2003). You can work work backward and
forward from there.  
\end{point}

\reply{
  Thanks for the additional suggestions!
  We've reworked the introduction, adding these (and many other) additional references
  into the (large and growing) literature. \revref
  We hope it's more representative now.
}

@@

\begin{point}{}
    Related to this: I find it non-intuitive to start the introduction with the
notion of ``nonidentifiability'', which is a consequence rather than a cause, the
cause being what you later call ``phenotypic equivalence'' and what had been
discussed under other names in other articles.
\end{point}

\reply{
  Good point. In the introdcution we now introduce nonidentifiability after discussing general cases of phenotypic equivalence. \revref
}

\begin{point}{}
    Still related: You give long lists of references in the introduction, but do
not provide the reader with any information about specific contributions. A bit
more would be helpful (starting with True and Haag coined the term
``developmental system drift'', etc).
\end{point}

\reply{
    We've tried to explain things better in the revised Introduction. \revref
    However, we find assigning specific contributions to be difficult
    because similar ideas appear in many disparate parts of the literature.
}

@@

\begin{point}{Eq. 3:}
    What does a coordinate change mean biologically? In our JC, we discussed for a while whether this is just a different parametrization of the exact same biological thing (like transformation to principle components). I think I now understand that it is more, but this should be explained better.
\end{point}

\reply{
    Biologically, a coordinate change may produce an unrecognizably different system
    (repression changed to activation or the like);
    we've added a hopefully clarifying sentence. \revref
}

\begin{point}{}
    ``Since gene networks can grow or shrink following gene duplications and deletions,
these additional degrees of freedom can apply in principle to any system.'' - are there examples of equivalent gene networks with a different size due to gene duplication or loss?
\end{point}

\reply{
    We're not aware of any biological examples of ``equivalent'' networks
    in more than one species at all,
    but we think it's not clear whether we understand the dynamics of \emph{any} gene regulatory network
    well enough to determine equivalence,
    so this is an absence of evidence, not evidence of absence.
    (There are plenty of examples of homologous networks with missing genes,
    but there's so many uncertainties we don't think bringing this up will clarify things.)
    % PLR: I don't think we need this, as the reviewers are aware of these examples.
    % Over larger evolutionary distances there is some evidence that structurally distinct
    % isofunctional networks of different size do indeed exist. \emph{E.g.}, the networks underlying circadian
    % rhythm in cyanobacteria and mammals, as discussed by \citet{sancar2008intelligent}.
    % It seems possible, in principle, that, on a smaller scale, such differences could exist between more closely related populations.

@@

\reply{
    We've removed the ``popgen'' part, but some of the comments below still apply.
}

\begin{point}{}
    For system drift, you assume that ``Selection will tend to restrain this
motion, but movement along the optimal set N is unconstrained, and so we expect
the population mean to drift along the optimal set like a particle diffusing.''
I see two major problems with this view. Both lead to slower divergence and
therefore run against your conclusions.
    \textit{(further points to follow)}
\end{point}

\reply{
    We have removed this part.
}

\begin{point}{(drift along a ridge)}
    In the presence of epistasis, evolution on a neutral network (a high-fitness
    ridge) is *not* due to drift alone, but also affected by (weak second order)
    selection in favor of mutational robustness / genetic canalization. In contrast
    to what you write, diffusion on the set of network coefficients corresponding
    to the optimal phenotype is not unbiased - even if the optimal phenotypes do
    indeed all have the same fitness. Instead, selection drives the population to
    ``thicker'' parts of the network where the mean fitness of the population
    (including a cloud of mutants) is higher that on a narrow ridge. This is the
    basis of the evolution of robustness/canalization. It is possible to account
    for this effect, see \citet{hermisson2003epistasis,alvarez2009effects}, 
    or also \citet{rice1998evolution,neutral_network_physics}.
    Note that epistasis is necessary for the neutral evolution of incompatibilities,
    which is what you are aiming for.
\end{point}

\reply{
    Thanks for pointing out this important omission.
    We have added mention of this (and these citations),
    including a paragraph to the discussion. \revref
}

\begin{point}{(drift along a ridge)}
    a second problem results from the fact that evolution on a high-fitness ridge
  often requires coordinated changes at many loci. Take the oscillator system
that you use as an example: simultaneous changes at two genes are required to
maintain the phenotype. This leads to a phenomenon called ``adaptive inertia''
(see \citet{baatz1997adaptive} and \citet{alvarez2009effects}),
which effectively slows down the movement along the ridge considerably. This
problem applies, in particular, in a ``house-of-cards'' mutation regime when
where rarely two mutations occur together on the same haplotype. In small
populations, it typically requires that populations drift through shallow
fitness valleys. While this is possible, it slows down the process.  
\end{point}

\reply{
    Again, this is an important point covered in additional discussion.
    \revref
}

@@

\begin{point}{(drift along a ridge)}
    The relevance of both effects could be studied by simulations in a simple
example (eg the oscillator that is used as an illustration in the ms anyway).
\end{point}

\reply{
    We agree, and hope to do this in our follow-up for this part of the paper.
}

\begin{point}{}
    If speciation due to accumulation of incompatibilities does not occur in
allopatry, but under (even weak) gene flow, some degree of positive selection
is always needed (\citet{bank2012limits} for a proof). This also
means that even weak gene flow will counteract the process described in the ms.
\end{point}

\reply{
    Also a good point - we've added mention of this as well. \revref
}
